Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-16 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
edited 4.2

\subsection{Theoretical evidence of linear proportionality between $m$, $c$ and $l$} \label{subsec_theoretic_linearity}

There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$. In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

{\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
proportional to 1/l which gives as that part, and the effect of c should also
be quite linear when c is not very close to 1 and m is not very small as it
linearly makes the interval shorter so that the there are less zero-level
upcrossings available in the basis functions. You can visualize this using
{\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
(ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
spectral densities and when they go to zero. This is illustrated in the same
notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
(linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
basis functions. These spectral densities don't depend on c. When c increaes
the basis functions are truncated and for example going from c=1.5 to c=2.25,
basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
functions are needed. So we have proportionality c/l. }

{\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}

>>>>

\subsection{Near linear proportionality between $m$, $c$, and $l$} \label{subsec_theoretic_linearity}

A priori the terms in the series expansion \eqref{eq_approxf_multi} with very small spectral density are unlikely to contribute to the approximation. Given the boundary factor $c$ and the length-scale $\ell$, we can compute cumulative sum of the spectral densities and find out how many basis functions are a priori explaining almost 100\% of the variation. Thus, given $c$ and $\ell$, we can estimate a good choice for the number of basis functions $m$ for any covariance function.

When considering exponentiated quadratic and Mat{\'e}rn covariance functions, we can show with simple algebra that when when $c$ is larger than the minimal value recommendation (see TODO ADD REF HERE), the $m$ first terms needed to explain almost 100\% of the variation has near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$ and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, the smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$, the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$ if $\ell$ or $c$ is changed.

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

% {\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}
